The Correspondence Analysis of Two Mode Networks Revisited

### 2025-03-25 main-R1.tex

\subsection{Organization of the Paper} \label{subsec:org}
The rest of the paper is organized as follows. In Section~\ref{sec:prevuses}, I review previous SNA work dealing with the uses of CA to analyze two-mode network data. Then, in Section~\ref{sec:ref2mode}, I consider a recent re-invention of CA (the economic complexity index) used to produce a type of Bonacich-style dual scoring via iterative scoring across the two modes of the affiliation matrix. This leads naturally to an abbreviated way of computing the same scores---CA and Bonacich---via spectral decomposition of the one-mode projection of the affiliation matrix for each set of nodes (degree-weighted in the case of CA and unweighted in the case of Bonacich). Section~\ref{sec:comparing} uses the classic Southern Women dataset to show the payoff of this approach, focusing on the distinct substantive insights extracted by CA and Bonacich scoring in a venerable dataset, showing that the CA-blocked affiliation matrix reveals a dual block partition between the two node sets (similar to that recovered by previous analyses of the same data using other methods), while the Bonacich-blocked affiliation recovers a core-periphery structure instead. Section~\ref{sec:cagensim} considers the hanging question of the exact substantive interpretation of the node groupings revealed by CA-style blocking, which are shown to be fairly close to those obtained by a computationally unrelated approach based on the idea of ``generalized relational similarity,'' suggesting that the CA blocks obey a logic distinct from structural equivalence. Finally, Section~\ref{sec:disc} closes with a summary of key points in the argument and suggestions for future development of the linkages between CA and positional and centrality analysis in two-mode networks.

@@

\begin{figure}[ht!]
    \captionsetup[subfigure]{font=footnotesize,labelfont=footnotesize}
    \centering
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/ca-reord1.png}
            \caption{}
            \label{fig:ca-reord1}
    \end{subfigure}
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/bon-reord1.png}
            \caption{}
            \label{fig:bon-reord1}
    \end{subfigure}
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/bon-reord2.png}
            \caption{}
            \label{fig:bon-reord2}
    \end{subfigure}
    \caption{Southern Women Affiliation matrix with re-ordered rows and columns: (a) row and columns re-ordered according to the person and group score on the first CA dimension, (b) row and columns re-ordered according to person and group Eigenvector centrality score, (c) row and columns re-ordered according to person and group score on the second Bonacich dimension.}
    \label{fig:ca-v-bon}
\end{figure}

@@

Note that equations~\ref{eq:simrank1} and~\ref{eq:simrank2} share a formal similarity with HH's ``method of reflections'' discussed earlier, in particular, the fact that both compute quantities based on nodes in the other mode averaged by the degree of nodes in the focal mode. The key difference is that SimRank works directly with pairwise comparisons between node dyads \citep{jeh2002simrank}. Nevertheless, this double weighting by degree should make us suspect that the results of the SimRank similarity analysis would not be too far from those obtained via CA, given the mathematical equivalence of CA and the method of reflections \citep{mealy2019interpreting, van2021correspondence}.

And indeed, they are not. Figure~\ref{fig:simrank}(a) shows the original affiliation matrix, this time with rows and columns re-arranged according to the values of the main non-trivial eigenvector (corresponding to the second-largest eigenvalue) of the SimRank similarity matrices---with $\alpha$ set to $0.8$---for the row and column nodes. Like in Figure~\ref{fig:ca-v-bon}(a), this reshuffling reveals the block-diagonal structure separating the two communities of persons and groups revealed by CA, suggesting that CA and SimRank uncover a similar underlying partitioning of the two-mode network's community structure.

Figure~\ref{fig:simrank}(b) and Figure~\ref{fig:simrank}(d) show a plot similar to that shown in Figures~\ref{fig:ca-eigvec}(a) and Figure~\ref{fig:ca-eigvec}(b) but this time with the main eigenvector of SimRank similarity matrices of both persons and groups on the x-axis and the corresponding rank on the y-axis. Looking at the plots from left to right, we can see that the partitioning of the node sets on the first informative eigenvector is substantively equivalent to those revealed by the first CA eigenvector, separating similar blocks of people and events.

Figure~\ref{fig:simrank}(f) shows the correspondence plot of the first and second eigenvectors of the SimRank similarity matrices for both persons and groups. Once again, this plot is substantively equivalent (in terms of the internode groupings and distances) to that shown in Figure~\ref{fig:ca-sim}(c), suggesting that the first-CA dimensions of the two-mode network recover clusters of ``similar'' entities in the two node sets, where internode similarity is a generalized relational similarity as defined earlier. Thus, [$Flora, Olivia$] appear close because they choose similar groups, as do [$Sylvia, Katherine, Helen$]. In the same way, groups [$E9, E11$] are close in the plot because they are chosen by similar people, as are events [$E10, E12, E13, E14$]. Figures~\ref{fig:simrank}(g) and (h) show a regression plot of the first CA eigenvector for persons (top) and groups (bottom) on the x-axis and the first eigenvector of the SimRank similarity matrix for both persons and groups on the y-axis. As we can see, the CA and SimRank ordering of nodes along the first dimension agree quite closely ($r = 0.98$ for persons and $r = 0.99$ for groups), confirming that these first eigenvectors capture the same (GSR) information across the two approaches.

>>>>

\subsection{Organization of the Paper} \label{subsec:org}
The rest of the paper is organized as follows. In Section~\ref{sec:prevuses}, I review previous SNA work dealing with the uses of CA to analyze two-mode network data. Then, in Section~\ref{sec:ref2mode}, I consider a recent re-invention of CA (the economic complexity index) used to produce a type of Bonacich-style dual scoring via iterative scoring across the two modes of the affiliation matrix. This leads naturally to an abbreviated way of computing the same scores---CA and Bonacich---via spectral decomposition of the one-mode projection of the affiliation matrix for each set of nodes (degree-weighted in the case of CA and unweighted in the case of Bonacich). Section~\ref{sec:comparing} uses the classic Southern Women dataset to show the payoff of this approach, focusing on the distinct substantive insights extracted by CA and Bonacich scoring in a venerable dataset, showing that the CA-blocked affiliation matrix reveals a dual block partition between the two node sets (similar to that recovered by previous analyses of the same data using other methods), while the Bonacich-blocked affiliation recovers a core-periphery structure instead. In addition I show that considering higher dimensions of the Bonacich-style decomposition allows us to discover secondary core-periphery structures in the two-mode network, while also leading to a meaningful clustering of the nodes based on structural equivalence. Section~\ref{sec:cagensim} considers the hanging question of the exact substantive interpretation of the node clusters revealed by CA-style blocking, which are shown to be fairly close to those obtained by a computationally unrelated approach based on the idea of ``generalized relational similarity,'' suggesting that the CA blocks are distinct from those obtained following the logic of structural equivalence. Finally, Section~\ref{sec:disc} closes with a summary of key points in the argument and suggestions for future development of the linkages between CA and positional and centrality analysis in two-mode networks.

@@

\begin{figure}[ht!]
    \captionsetup[subfigure]{font=footnotesize,labelfont=footnotesize}
    \centering
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/ca-reord1.png}
            \caption{First CA Dimension}
            \label{fig:ca-reord1}
    \end{subfigure}
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/bon-reord1.png}
            \caption{First Bonacich Dimension}
            \label{fig:bon-reord1}
    \end{subfigure}
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/bon-reord2.png}
            \caption{Second Bonacich Dimension}
            \label{fig:bon-reord2}
    \end{subfigure}
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/SR-reord1.png}
            \caption{Second SimRank Eigenvector}
            \label{fig:sr-reord1}
    \end{subfigure}
    \caption{Southern Women Affiliation matrix with re-ordered rows and columns: (a) rows and columns re-ordered according to the person and group score on the first CA dimension, (b) rows and columns re-ordered according to person and group Eigenvector centrality score, (c) rows and columns re-ordered according to person and group score on the second Bonacich dimension, (d) rows and columns re-ordered according to the second eigenvector of the SimRank similarity matrix.}
    \label{fig:ca-v-bon}
\end{figure}

@@

Note that equations~\ref{eq:simrank1} and~\ref{eq:simrank2} share a formal similarity with HH's ``method of reflections'' discussed earlier, in particular, the fact that both compute quantities based on nodes in the other mode averaged by the degree of nodes in the focal mode. The key difference is that SimRank works directly with pairwise comparisons between node dyads \citep{jeh2002simrank}. Nevertheless, this double weighting by degree should make us suspect that the results of the SimRank similarity analysis would not be too far from those obtained via CA, given the mathematical equivalence of CA and the method of reflections \citep{mealy2019interpreting, van2021correspondence}. Beyond that, the SimRank approach can be thought of as yet another way of obtaining one-mode projections from an original two-mode network, in line with Everett and Borgatti's ``dual projection'' approach \citep{everett2013dual}.

Figure~\ref{fig:sr-reord1} shows the original affiliation matrix, this time with rows and columns re-arranged according to the values of the main non-trivial eigenvector (corresponding to the second-largest eigenvalue) of the SimRank similarity matrices---with $\alpha$ set to $0.8$---for the row and column nodes. Like in Figure~\ref{fig:ca-v-bon}(a), this reshuffling displays a block-diagonal structure separating the two communities of persons and groups as that revealed by CA, suggesting that CA and SimRank uncover a similar underlying partitioning of the two-mode network's community structure. Figure~\ref{fig:simrank}(a) and Figure~\ref{fig:simrank}(b) show a plot similar to that shown in Figures~\ref{fig:ca-eigvec}(a) and Figure~\ref{fig:ca-eigvec}(b) but this time with the main eigenvector of SimRank similarity matrices of both persons and groups on the x-axis and the corresponding rank on the y-axis. Looking at the plots from left to right, we can see that the partitioning of the node sets on the most informative eigenvector of the SimRank similarity matrix is substantively equivalent to those revealed by the first CA eigenvector, separating similar blocks of people and events.

Figure~\ref{fig:simrank}(f) shows the correspondence plot of the two main non-trivial eigenvectors of the SimRank similarity matrices (corresponding to the second and third eigevectors) for both persons and groups. Once again, this plot is substantively equivalent (in terms of the internode groupings and distances) to that shown in Figure~\ref{fig:ca-sim}(c), suggesting that the first-CA dimensions of the two-mode network recover clusters of ``similar'' entities in the two node sets, where internode similarity is a generalized relational similarity. Thus, [$Flora, Olivia$] appear close because they choose similar groups, as do [$Sylvia, Katherine, Helen$]. In the same way, groups [$E9, E11$] are close in the plot because they are chosen by similar people, as are events [$E10, E12, E13, E14$]. Figures~\ref{fig:simrank}(g) and (h) show a regression plot of the first CA eigenvector for persons (top) and groups (bottom) on the x-axis and the first eigenvector of the SimRank similarity matrix for both persons and groups on the y-axis. As we can see, the CA and SimRank ordering of nodes along the first dimension agree quite closely ($r = 0.98$ for persons and $r = 0.99$ for groups), confirming that these first eigenvectors capture the same (GSR) information across the two approaches.
